% ============================================================================
%  ABSTRACT  ->  replaces \lipsum[1] and \lipsum[2][1-5] under the
%  \frontmatterheading{ABSTRACT} block in main.tex (~line 213)
% ============================================================================

Every year, millions of salaried Indians file their income tax returns while
sitting on an awkward information gap: the Income Tax Department already knows
about most of their financial activity through the Annual Information Statement
(AIS), yet the tools people use to file still work almost entirely from
self-reported numbers. When the two views disagree, the taxpayer is the last to
find out -- usually through a Section 143(1) notice months later. This thesis
presents FinITR-AI, a locally-runnable, multi-agent system that closes that gap
by reconciling a taxpayer's Form 16, bank statement and AIS before the return is
ever filed.

FinITR-AI is built as a ReAct-style pipeline of four specialised agents -- an
Auditor that cross-checks the documents and scores notice risk, an Optimizer
that compares the Old and New tax regimes, a Compliance agent that selects the
correct ITR form and maps income to schedules, and a Critic that verifies every
claim against the source documents and forces a re-run when something cannot be
grounded. A key design decision is that no arithmetic is ever left to the
language model: all tax computation flows through a deterministic engine that
hard-codes the FY 2025-26 slabs, surcharge, cess, rebate and marginal-relief
rules.

Supporting the pipeline are two trained models --- a three-stage transaction
classifier that labels every bank-statement line, and a logistic-regression
notice predictor tuned for zero false negatives --- together with an NLI-based
faithfulness verifier and a structured retriever over a corpus of tax-law
sections. We evaluate the system on IndianTaxBench, a purpose-built suite of 140
labelled cases spanning salary, regime comparison, capital gains, crypto, AIS
reconciliation, ITR-form selection, adversarial traps, and CTC restructuring. On
the 100-case suite FinITR-AI reaches 94.0\% tax-computation accuracy with a 6.8\%
hallucination rate, against 70--74\% accuracy and 24--27\% hallucination for
prompted GPT-4o and Gemini 2.0 Flash, and it passes all sixteen prompts of a
direct manual comparison on which the baselines each fail two. An ablation
isolates the deterministic calculator and the critic loop as the two
load-bearing components, the former worth roughly forty points of tax accuracy
and the latter responsible for keeping hallucination low. The result is a
private, explainable filing assistant that combines the arithmetic reliability
of rule-based software with the flexible reasoning of large language models.
