% ============================================================================
%  CHAPTER 7: CONCLUSION AND FUTURE SCOPE   (expanded version)
%  Replaces the entire Chapter 7 body in main.tex.
% ============================================================================

\chapter{Conclusion and Future Scope}
\label{chap:conclusion}

This project began with a problem that most Indian taxpayers live with without
ever naming it. They are asked to declare their own income, but they are
measured against a far more complete record --- the AIS --- that the tools they
file with never actually consult, and when the two disagree, the entire cost of
the disagreement lands on them. FinITR-AI was built to close that gap before it
becomes a notice. It reconciles Form 16, the bank statement, and the AIS while
the return is still being prepared; it computes the exact liability under both
FY 2025-26 regimes through a deterministic engine; it selects the correct ITR
form and maps each income source to its schedule; and it verifies every claim it
makes against the source documents and a structured corpus of tax law before
letting that claim into the final report.

If there is a single lesson the work teaches, it is architectural rather than
about any particular model. The way to make a language model trustworthy for
taxes is not to find a larger model or to write a more elaborate prompt; it is to
take the arithmetic away from the model entirely and place it behind a
deterministic engine, and then to put a critic in the loop with the authority to
reject anything the documents do not support. The evaluation bears this out. On
the IndianTaxBench suite the system reaches 94.0\% tax-computation accuracy with
a 6.8\% hallucination rate, against 70 to 74\% accuracy and 24 to 27\%
hallucination for prompted GPT-4o and Gemini 2.0 Flash, and the ablation
attributes those gains specifically to the calculator and the critic rather than
to anything diffuse. Just as importantly, the whole pipeline runs offline, so
the most sensitive documents a person owns never leave their own machine --- a
property the cloud-based alternatives cannot offer at any level of accuracy.

The system is not finished, and several of its limitations point cleanly to the
next pieces of work.

\noindent\textbf{Future Enhancements:}

\begin{itemize}
    \item \textbf{Close the Schedule OS recall gap.} A conservative
          income-bracket heuristic --- inferring Schedule OS when a taxpayer
          holds a savings account and earns above a threshold --- would recover
          the roughly thirteen percent of interest-income cases the system
          currently leaves out, and the precision cost of doing so should be
          measured rather than assumed.
    \item \textbf{Add a noise-correction stage before the engine.} The held-out
          degradation comes from OCR rounding flowing straight into the
          deterministic calculator. A dedicated input-cleaning module that
          cross-checks parsed figures across the available documents would
          recover much of that loss before it reaches the arithmetic.
    \item \textbf{Refine multi-signal risk escalation.} The risk scorer should
          be strengthened so that a single severe signal --- undeclared crypto
          visible only as a bank deposit, for example --- can escalate to
          CRITICAL without waiting for a second corroborating signal, which is
          the precise cause of the one notable misclassification in the results.
    \item \textbf{Broaden coverage beyond the salaried filer.} Expanding the
          training and validation cases for business and presumptive income
          (ITR-3 and ITR-4) would address the multi-income edge cases behind the
          eight-point form-selection gap and widen the system's safe operating
          range.
    \item \textbf{Integrate direct filing.} Connecting the validated ITR-2 JSON
          export to the Income Tax Department's upload schema would let the tool
          carry a user from raw documents to a filed return end to end, instead
          of stopping at a filing-ready report.
\end{itemize}

Even in its present form, FinITR-AI demonstrates something worth stating plainly:
a private, explainable, locally-runnable assistant can match the arithmetic
reliability of established rule-based software while adding the AIS
reconciliation, the notice-risk modelling, and the natural-language reasoning
those tools lack --- and it can do all of this without ever sending a taxpayer's
documents to a server it does not control. The broader principle generalises
beyond tax. For any high-stakes, rule-governed domain, the pattern that worked
here is likely to work again: let the language model reason and explain, force it
to defer every consequential computation to a verified tool, and keep a critic in
the loop with the standing to refuse whatever cannot be grounded.
