% ============================================================================
%  BIBLIOGRAPHY
%  Replaces the dummy \bibitem entries inside the existing
%  \begin{thebibliography}{99} ... \end{thebibliography} block in main.tex.
%  Keys (ref1..ref12) match the \cite{} calls used across the chapters.
%
%  NOTE: Verify each URL and access date before final submission. The years
%  reflect the FY 2025-26 (AY 2026-27) context of this project.
% ============================================================================

\bibitem{ref1}
Income Tax Department, Government of India. (2024). \textit{e-Filing Portal and
Offline Utilities for Income Tax Returns}. Central Board of Direct Taxes.
\url{https://www.incometax.gov.in/}

\bibitem{ref2}
ClearTax (Defmacro Software Pvt. Ltd.). (2024). \textit{ClearTax Income Tax
e-Filing Platform}. \url{https://cleartax.in/}

\bibitem{ref3}
Quicko Infosoft Pvt. Ltd. (2024). \textit{Quicko -- Income Tax Filing and Tax
Planning Platform}. \url{https://quicko.com/}

\bibitem{ref4}
OpenAI. (2024). \textit{GPT-4o System Card}. OpenAI.
\url{https://openai.com/index/gpt-4o-system-card/}

\bibitem{ref5}
Google DeepMind. (2024). \textit{Gemini 2.0 Flash: Model and Technical
Overview}. Google.
\url{https://deepmind.google/technologies/gemini/}

\bibitem{ref6}
Wu, S., Irsoy, O., Lu, S., Dabravolski, V., Dredze, M., Gehrmann, S., Kambadur,
P., Rosenberg, D., \& Mann, G. (2023). \textit{BloombergGPT: A Large Language
Model for Finance}. arXiv:2303.17564.

\bibitem{ref7}
Yao, S., Zhao, J., Yu, D., Du, N., Shafran, I., Narasimhan, K., \& Cao, Y.
(2023). \textit{ReAct: Synergizing Reasoning and Acting in Language Models}.
International Conference on Learning Representations (ICLR). arXiv:2210.03629.

\bibitem{ref8}
Lewis, P., Perez, E., Piktus, A., Petroni, F., Karpukhin, V., Goyal, N., et al.
(2020). \textit{Retrieval-Augmented Generation for Knowledge-Intensive NLP
Tasks}. Advances in Neural Information Processing Systems (NeurIPS).
arXiv:2005.11401.

\bibitem{ref9}
He, P., Liu, X., Gao, J., \& Chen, W. (2021). \textit{DeBERTa:
Decoding-enhanced BERT with Disentangled Attention}. International Conference on
Learning Representations (ICLR). arXiv:2006.03654.

\bibitem{ref10}
Reimers, N., \& Gurevych, I. (2019). \textit{Sentence-BERT: Sentence Embeddings
using Siamese BERT-Networks}. Conference on Empirical Methods in Natural Language
Processing (EMNLP). arXiv:1908.10084.

\bibitem{ref11}
Pedregosa, F., Varoquaux, G., Gramfort, A., Michel, V., Thirion, B., Grisel, O.,
et al. (2011). \textit{Scikit-learn: Machine Learning in Python}. Journal of
Machine Learning Research, 12, 2825--2830.

\bibitem{ref12}
Central Board of Direct Taxes, Ministry of Finance, Government of India. (2024).
\textit{Annual Report 2023-24}. \url{https://www.incometax.gov.in/}
